---
title: "BeginAnalysis"
author: "Eamon"
date: "2024-12-18"
output: html_document
---
# Initialisation
## Set Up
### Load Libraries
```{r}
library(tidyverse)
library(corrplot) # For correlation plots
library(FactoMineR) # For PCA
library(factoextra) # For PCA plots
library(vegan) # For CCA
library(ggplot2)
```

### 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Clear environment
```{r}
rm(list = ls())
```


### Specify the FilePath and Read CSV
```{r}
path <- "/Users/eamon/Desktop/MathToolsProject/R_Project/MathematicalTools/lizard.csv"

# Read a CSV file
data <- read.csv(path)

# View the first few rows of the data
head(data)
```

## Cleaning

### Change Variables Types
```{r}
# Check the type of each column
sapply(data, class)

# Convert SD.Female.adult.weight..g. to numeric
data$SD.Female.adult.weight..g. <- as.numeric(data$SD.Female.adult.weight..g.)

# Convert categorical columns to factors
data$Species <- as.factor(data$Species)
data$Genus <- as.factor(data$Genus)
data$Family<- as.factor(data$Family)
data$Mode.of.reproduction <- as.factor(data$Mode.of.reproduction)
data$Foraging.Mode <- as.factor(data$Foraging.Mode)
data$Distribution <- as.factor(data$Distribution)
data$Prefered.Habitat.Type <- as.factor(data$Prefered.Habitat.Type)
```

### Check for and Remove Duplicates
```{r}
# Check for Duplicates
duplicates <- data[duplicated(data), ]
print("Duplicate Rows:")
print(duplicates)

# Remove duplicates
data <- data[!duplicated(data), ]
```

## NAs
### Check Amount of Data Loss If All NAs Removed
Here it was found that if all rows containing an NA are removed the dataset is reduced from 734 rows to 146 rows, which is too much loss.
```{r}
data_clean <- na.omit(data)
nrow(data)
nrow(data_clean)
```

### Calculate Percentage of Each Variable Which Are NA
The aim here is to understand the percentage of each variable which are NA so we can choose which to focus on in our analysis.
```{r}
# Calculate percentage of NA values for each column
na_percentage <- colSums(is.na(data)) / nrow(data) * 100

# Convert NA percentages to a data frame
na_df <- data.frame(
  Column = names(na_percentage),
  NA_Percentage = na_percentage
)

# Create bar plot
ggplot(na_df, aes(x = Column, y = NA_Percentage)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(
    title = "Percentage of NA Values per Column",
    x = "Columns",
    y = "Percentage (%)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1) # Rotate x-axis labels
  )
```


# Visualisation
## Correlations
### CorrPlot
```{r}
data %>% 
  select_if(is.numeric) %>% # Select only numeric columns
  cor(use="pairwise.complete.obs") %>% # Calculate the empirical correlation matrix
  corrplot() # Then graph this matrix
```



