---
title: "BeginAnalysis"
author: "Eamon"
date: "2024-12-18"
output: html_document
---
# Initialisation
## Set Up
### Load Libraries
```{r}
library(tidyverse)
library(corrplot) # For correlation plots
library(FactoMineR) # For PCA
library(factoextra) # For PCA plots
library(vegan) # For CCA
library(ggplot2)
library(rsample) # For sampling of data (split, analysis,..)
library(rpart)
library(rpart.plot)
library(randomForest)
```

### 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Clear environment
```{r}
rm(list = ls())
```

### Specify the FilePath and Read CSV
```{r}
path <- "/Users/eamon/Desktop/MathToolsProject/R_Project/MathematicalTools/lizard.csv"

# Read a CSV file
data <- read.csv(path)
```
### Column Names for Copying
```{r}
str(data)
```
### Comparison of Similar Variables
```{r}
# Are clutch frequency and clutches per year the same?  
cor(data$Clutch.Frequency, data$Clutches.per.year, use = "complete.obs")
plot(data$Clutch.Frequency~data$Clutches.per.year)
unique(data$Clutch.Frequency)
unique(data$Clutches.per.year)

# Are "F.SVL.at.maturity..mm." and "Mean.F.SVL.adults..mm." the same?
cor(data$F.SVL.at.maturity..mm., data$Mean.F.SVL.adults..mm., use="complete.obs")
plot(data$F.SVL.at.maturity..mm.~data$Mean.F.SVL.adults..mm.)
```
Seems like Clutch Frequency and Clutches per Year are highly related, but Clutches Per Year goes up to 6 and Clutch Frequency only ranges between 1 and 2.
Seems like "F.SVL.at.maturity..mm." and "Mean.F.SVL.adults..mm." are the same.
Therefore one of each pair should be excluded.


##Initialise Settings of the Analysis
```{r}
# These variables are kept as required to answer the research questions.
# This will result in the "data_filtered" and "data_clean" dataframes.
variables_to_keep <- c(
  ### Categorical Variables
  #"Species",
  #"Genus",
  "Family",
  #"Population",
  "Mode.of.reproduction",
  #"Source",
  "Distribution",
  "Prefered.Habitat.Type",
  "Foraging.mode",
  
  ### Location
  #"Longitude",
  #"Latitude",
  
  ### Numerical Variables
  "Average.Female.adult.weight..g.",
  "Mean.F.SVL.adults..mm.",
  #"F.SVL.at.maturity..mm.", # Removed as "Mean.F.SVL.adults..mm." is almost identical.
  "Offspring.SVL..mm.",
  "Mean.Clutch.Size",
  #"Clutches.per.year",
  "Clutch.Frequency",
  "RCM"
  
  ### Standard Deviations and Sample Sizes
  #"SD.F.SVL.adults..mm.",
  #"Sample.size.Mean.F.SVL.adults",
  #"SD.Female.adult.weight..g.",
  #"Sample.Size.Female.adult.weight"
)

# Specify the variables to be included in the PCA
# This will be result in the "data_scaled" dataframe
variables_to_use_PCA <- c(
  ### Location
  #"Longitude",
  #"Latitude",
  
  ### Means and Numerical
  "Average.Female.adult.weight..g.",
  "Mean.F.SVL.adults..mm.",
  "F.SVL.at.maturity..mm.",
  "Offspring.SVL..mm.",
  "Mean.Clutch.Size"
  #"Clutches.per.year",
  #"Clutch.Frequency",
  #"RCM"
  
  ### Standard Deviations and Sample Sizes
  #"SD.Female.adult.weight..g.",
  #"SD.F.SVL.adults..mm.",
  #"Sample.Size.Female.adult.weight",
  #"Sample.size.Mean.F.SVL.adults",
)

# Specify the Variables to Use in the Random Forests Analysis
variables_to_use_RF <- c(
  ### Categorical Variables
  #"Species",
  #"Genus",
  #"Family",
  #"Population",
  "Mode.of.reproduction",
  #"Source",
  "Distribution",
  "Prefered.Habitat.Type",
  "Foraging.Mode",
  "RCM",
  
  ### Location
  #"Longitude",
  #"Latitude",
  
  ### Numerical Variables
  "Average.Female.adult.weight..g.",
  "Mean.F.SVL.adults..mm.",
  #"F.SVL.at.maturity..mm.",
  "Offspring.SVL..mm.",
  "Mean.Clutch.Size",
  #"Clutches.per.year"
  "Clutch.Frequency"
  #"RCM"
  
  ### Standard Deviations and Sample Sizes
  #"SD.F.SVL.adults..mm.",
  #"Sample.size.Mean.F.SVL.adults",
  #"SD.Female.adult.weight..g.",
  #"Sample.Size.Female.adult.weight"
)
```

## Cleaning
### Change Variables Types
```{r}
# Check the type of each column
sapply(data, class)

# Convert categorical columns to factors
data$Species <- as.factor(data$Species)
data$Genus <- as.factor(data$Genus)
data$Family<- as.factor(data$Family)
data$Mode.of.reproduction <- as.factor(data$Mode.of.reproduction)
data$Foraging.Mode <- as.factor(data$Foraging.Mode)
data$Distribution <- as.factor(data$Distribution)
data$Prefered.Habitat.Type <- as.factor(data$Prefered.Habitat.Type)
```

### Check for and Remove Duplicates
```{r}
# Check for Duplicates
duplicates <- data[duplicated(data), ]
print("Duplicate Rows:")
print(duplicates)

# Remove duplicates
data <- data[!duplicated(data), ]
```

### Calculate Percentage of Each Variable Which Are NA
```{r}
# Replace empty cells with NA for factors
data[data == ''] <- NA

# Calculate percentage of NA values for each column
na_percentage <- colSums(is.na(data)) / nrow(data) * 100

# Convert NA percentages to a data frame
na_df <- data.frame(
  Column = names(na_percentage),
  NA_Percentage = na_percentage
)

# Create bar plot
ggplot(na_df, aes(x = Column, y = NA_Percentage)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(
    title = "Percentage of NA Values per Column",
    x = "Columns",
    y = "Percentage (%)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1) # Rotate x-axis labels
  )
```

### Remove Unneeded Variables
Here we removed the variables that we didn't plan to use in any part of the analysis. 
This was done before filtering out NAs to avoid unneccessary data loss.
All rows of the RCM column which are NA are also NA for mean female adult weight (which makes sense as they're related).
```{r}
# Select only the desired variables
data_filtered <- data %>%
  select(any_of(variables_to_keep))
```

## NAs
### Check Amount of Data Loss If All NAs Removed
```{r}
data_clean <- na.omit(data_filtered)
nrow(data)
# Too many rows lost if all NAs removed
nrow(data_clean)
```

# PCA
### Create New Dataset retaining only the PCA variables
```{r}
data_PCA <- data %>%
  select(any_of(variables_to_use_PCA)) %>%
  na.omit()
```

### Scale the Data
```{r}
data_PCA <- data_PCA %>% 
  #Only numeric columns are selected
  mutate_all(.funs = scale)
```

### Visualise with a Correlation
```{r}
### CorrPlot
data_PCA %>% 
  cor(use="pairwise.complete.obs") %>% # Calculate the empirical correlation matrix
  corrplot() # Then graph this matrix
```


### Run the PCA
```{r}
result_pca <- PCA(data_PCA, 
               scale.unit = TRUE, # Option to center and scale data (useless here)
               ncp = 18, # Number of components to keep (here, all)
               graph = FALSE)
```

### Result of PCA
```{r}
names(result_pca) # It's a list
result_pca
```

### Percentage of Information Retained
```{r}
fviz_eig(result_pca, choice = "variance")
```
### Correlation between Each Variable and the Principle Components
```{r}
corrplot(result_pca$var$cor)
```



### Visualise the PCA
```{r}
# Representation in the first principal plane
fviz_pca_var(result_pca,
             axes = c(1, 2)) # Number of axes to represent 
fviz_pca_var(result_pca,
             axes = c(1, 3)) # Number of axes to represent 
fviz_pca_var(result_pca,
             axes = c(2, 3)) # Number of axes to represent 
fviz_pca_var(result_pca,
             axes = c(1, 4))
```

### Represent Individuals
```{r}
# Representation in the first principal plane
fviz_pca_ind(result_pca,
             axes = c(1, 2))
fviz_pca_ind(result_pca,
             axes = c(1, 3))
fviz_pca_ind(result_pca,
             axes = c(2, 3))
fviz_pca_ind(result_pca,
             axes = c(3, 4))
```
### Plot the Individuals with the Variable Arrows and the Categorical Coloring
```{r}
# Representation in the first principal plane
fviz_pca_biplot(result_pca,
             axes = c(1, 2),
             col.ind = data_clean$Distribution)
fviz_pca_biplot(result_pca,
             axes = c(1, 3),
             col.ind = data_clean$Distribution)
fviz_pca_biplot(result_pca,
             axes = c(2, 3),
             col.ind = data_clean$Distribution)
fviz_pca_biplot(result_pca,
             axes = c(3, 4),
             col.ind = data_clean$Distribution)
```
# K-Means Clustering
### 
```{r}
# Create Clusters For Four Groups
first_kmeans <- data_PCA %>% 
  kmeans(centers = 4)

# Within Group Inertia
first_kmeans$tot.withinss

kmeans_5groups<- data_PCA %>%
  kmeans(centers = 5)

kmeans_4groups<- data_PCA %>%
  kmeans(centers = 4)

kmeans_3groups<- data_PCA %>%
  kmeans(centers=3)

kmeans_2groups <- data_PCA %>%
  kmeans(centers=2)

# Extract the Clustering Groups
clusters_5groups <- kmeans_5groups$cluster
clusters_4groups <- kmeans_4groups$cluster
clusters_3groups <- kmeans_3groups$cluster
clusters_2groups <- kmeans_2groups$cluster
```

## Plot Clusters 
```{r}
# Representation in the first principal plane
fviz_pca_biplot(result_pca,
             axes = c(1, 2),
             col.ind = clusters_4groups,
             fill.var = "Red")
fviz_pca_biplot(result_pca,
             axes = c(1, 3),
             col.ind = clusters_4groups)
fviz_pca_biplot(result_pca,
             axes = c(2, 3),
             col.ind = clusters_4groups)
fviz_pca_biplot(result_pca,
             axes = c(3, 4),
             col.ind = clusters_4groups)
```


# Unsupervised Learning
##Split Data
###Filter Data Variables
```{r}
data_RF <- data_clean %>%
  select(any_of(variables_to_use_RF))
str(data_RF)
```

### Initial Split 
```{r}
split_data <- data_RF %>%
  initial_split(prop = 0.8, strata = "RCM")

# Train and Test Data
train_data <- analysis(split_data)
test_data <- assessment(split_data)
```

## Random Forests
### Create the RF
```{r}
forest <- randomForest(RCM ~ ., # Formula for prediction
                      data = train_data, # Data for training
                      ntree = 5000, # Number of trees
                      maxnodes = 10, # Number of maximum leaves for each tree
                      mtry = 4, # Number of variables for each tree
                      importance = TRUE) # Computation of importance
```

### Test the Accuracy
```{r}
# Fast way of computing accuracy with the pipe operator %>% 
predict(forest, newdata = select(test_data, - RCM)) %>% 
  table(prediction = ., truth = test_data$RCM) %>% 
  {sum(diag(.)) / sum(.)}
```


