---
title: "BeginAnalysis"
author: "Eamon"
date: "2024-12-18"
output: html_document
---
# Initialisation
## Set Up
### Load Libraries
```{r}
library(tidyverse)
library(corrplot) # For correlation plots
library(FactoMineR) # For PCA
library(factoextra) # For PCA plots
library(vegan) # For CCA
library(ggplot2)
library(rsample) # For sampling of data (split, analysis,..)
library(rpart)
library(rpart.plot)
library(randomForest)
library(tibble)
library(tidyr)
library(gridExtra)
library(factoextra)
library(caret)
```

### 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Clear environment
```{r}
rm(list = ls())
```

### Specify the FilePath and Read CSV
```{r}
path <- "/Users/eamon/Desktop/MathToolsProject/R_Project/MathematicalTools/lizard.csv"

# Read a CSV file
data <- read.csv(path)
```
### Column Names for Copying
```{r}
str(data)
```
### Comparison of Similar Variables
```{r}
# Are clutch frequency and clutches per year the same?  
cor(data$Clutch.Frequency, data$Clutches.per.year, use = "complete.obs")
plot(data$Clutch.Frequency~data$Clutches.per.year)
unique(data$Clutch.Frequency)
unique(data$Clutches.per.year)

# Are "F.SVL.at.maturity..mm." and "Mean.F.SVL.adults..mm." the same?
cor(data$F.SVL.at.maturity..mm., data$Mean.F.SVL.adults..mm., use="complete.obs")
plot(data$F.SVL.at.maturity..mm.~data$Mean.F.SVL.adults..mm.)
```
Seems like Clutch Frequency and Clutches per Year are highly related, but Clutches Per Year goes up to 6 and Clutch Frequency only ranges between 1 and 2.
Seems like "F.SVL.at.maturity..mm." and "Mean.F.SVL.adults..mm." are the same.
Therefore one of each pair should be excluded.


##Initialise Settings of the Analysis
```{r}
# These variables are kept as required to answer the research questions.
# This will result in the "data_filtered" and "data_clean" dataframes.
variables_to_keep <- c(
  ### Categorical Variables
  #"Species",
  #"Genus",
  "Family",
  #"Population",
  "Mode.of.reproduction",
  #"Source",
  "Distribution",
  "Prefered.Habitat.Type",
  "Foraging.mode",
  
  ### Location
  #"Longitude",
  #"Latitude",
  
  ### Numerical Variables
  "Average.Female.adult.weight..g.",
  "Mean.F.SVL.adults..mm.",
  #"F.SVL.at.maturity..mm.", # Removed as "Mean.F.SVL.adults..mm." is almost identical.
  "Offspring.SVL..mm.",
  "Mean.Clutch.Size",
  "Clutches.per.year",
  #"Clutch.Frequency",
  "RCM"
  
  ### Standard Deviations and Sample Sizes
  #"SD.F.SVL.adults..mm.",
  #"Sample.size.Mean.F.SVL.adults",
  #"SD.Female.adult.weight..g.",
  #"Sample.Size.Female.adult.weight"
)

# Specify the variables to be included in the PCA
# This will be result in the "data_scaled" dataframe

variables_to_use_PCA <- c(
  ### Location
  #"Longitude",
  #"Latitude",
  
  ### Means and Numerical
  "Average.Female.adult.weight..g.",
  "Mean.F.SVL.adults..mm.",
  #"F.SVL.at.maturity..mm.",
  "Offspring.SVL..mm.",
  "Mean.Clutch.Size",
  "Clutches.per.year",
  #"Clutch.Frequency",
  "RCM"
  
  ### Standard Deviations and Sample Sizes
  #"SD.Female.adult.weight..g.",
  #"SD.F.SVL.adults..mm.",
  #"Sample.Size.Female.adult.weight",
  #"Sample.size.Mean.F.SVL.adults",
)

# Specify the Variables to Use in the Random Forests Analysis
variables_to_use_RF <- c(
  ### Categorical Variables
  #"Species",
  #"Genus",
  #"Family",
  #"Population",
  "Mode.of.reproduction",
  #"Source",
  "Distribution",
  "Prefered.Habitat.Type",
  "Foraging.Mode",
  "RCM",
  
  ### Location
  #"Longitude",
  #"Latitude",
  
  ### Numerical Variables
  "Average.Female.adult.weight..g.",
  "Mean.F.SVL.adults..mm.",
  "F.SVL.at.maturity..mm.",
  "Offspring.SVL..mm.",
  "Mean.Clutch.Size",
  "Clutches.per.year",
  "Clutch.Frequency",
  "RCM"
  
  ### Standard Deviations and Sample Sizes
  #"SD.F.SVL.adults..mm.",
  #"Sample.size.Mean.F.SVL.adults",
  #"SD.Female.adult.weight..g.",
  #"Sample.Size.Female.adult.weight"
)

variables_to_use_CCA <- c(
  #"Distribution",
  #"Foraging.Mode",
  #"Mode.of.reproduction",
  "Prefered.Habitat.Type",
  "Family",
  "Latitude"
)
```

## Cleaning
### Change Variable Types and Set Blanks as NA
```{r}
# Set numeric variables as numeric
data$SD.Female.adult.weight..g. <- as.numeric(data$SD.Female.adult.weight..g.)

str(data)

# Convert all blank values of categorical variables to NA
data<-data %>%
  mutate(across(where(is.character), ~ na_if(.x, ""))) %>%
  mutate(across(where(is.character), as.factor))

str(data)
```

### Check for and Remove Duplicates
```{r}
# Check for Duplicates
duplicates <- data[duplicated(data), ]
print("Duplicate Rows:")
print(duplicates)

# Remove duplicates
data <- data[!duplicated(data), ]
```

### Calculate Percentage of Each Variable Which Are NA
```{r}
# Replace empty cells with NA for factors
data[data == ''] <- NA

# Calculate percentage of NA values for each column
na_percentage <- colSums(is.na(data)) / nrow(data) * 100

# Convert NA percentages to a data frame
na_df <- data.frame(
  Column = names(na_percentage),
  NA_Percentage = na_percentage
)

# Create bar plot
ggplot(na_df, aes(x = Column, y = NA_Percentage)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(
    title = "Percentage of NA Values per Column",
    x = "Columns",
    y = "Percentage (%)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1) # Rotate x-axis labels
  )
```

### Remove Unneeded Variables
Here we removed the variables that we didn't plan to use in any part of the analysis. 
This was done before filtering out NAs to avoid unneccessary data loss.
All rows of the RCM column which are NA are also NA for mean female adult weight (which makes sense as they're related).
```{r}
# Select only the desired variables
data_filtered <- data %>%
  select(any_of(variables_to_keep))
``` 

## NAs
### Check Amount of Data Loss If All NAs Removed
```{r}
data_clean <- na.omit(data_filtered)
nrow(data)
# Too many rows lost if all NAs removed
nrow(data_clean)
```

# PCA
### Create New Dataset retaining only the PCA variables
```{r}
# Create new dataset by selecting only variables to use and removing rows with NAs
data_PCA <- data %>%
  select(any_of(variables_to_use_PCA)) %>%
  na.omit()

# Create ancillary dataset with all original variables but filtered to keep only rows present in the PCA dataset
# This will be used for coloration of the PCA graphs
data_PCA_all_vars <- data %>%
  filter(row_number() %in% rownames(data_PCA))
```

### Scale the Data
```{r}
data_PCA <- data_PCA %>% 
  #Only numeric columns are selected
  mutate_all(.funs = scale)
```

### Visualise with a Correlation
```{r}
### CorrPlot
data_PCA %>% 
  cor(use="pairwise.complete.obs") %>% # Calculate the empirical correlation matrix
  corrplot() # Then graph this matrix
```


### Run the PCA
```{r}
result_pca <- PCA(data_PCA, 
               scale.unit = TRUE, # Option to center and scale data (useless here)
               ncp = 18, # Number of components to keep (here, all)
               graph = FALSE)
```

### Result of PCA
```{r}
result_pca
result_pca$eig
result_pca$var
```

### Percentage of Information Retained
```{r}
fviz_eig(result_pca, choice = "variance")
```
### Correlation between Each Variable and the Principle Components
```{r}
corrplot(result_pca$var$cor)
```

### Visualise the PCA
```{r}
# Representation in the first principal plane
fviz_pca_var(result_pca,
             axes = c(1, 2)) # Number of axes to represent 
fviz_pca_var(result_pca,
             axes = c(1, 3)) # Number of axes to represent 
fviz_pca_var(result_pca,
             axes = c(2, 3)) # Number of axes to represent 
fviz_pca_var(result_pca,
             axes = c(1, 4))
```
### Plot the Individuals with the Variable Arrows and the Categorical Coloring
```{r}
Color_label <- "Distribution"

# Plot on Seperate Plots
fviz_pca_biplot(result_pca,
             axes = c(1, 2),
             col.ind = data_PCA_all_vars[[Color_label]])
fviz_pca_biplot(result_pca,
             axes = c(1, 3),
             col.ind = data_PCA_all_vars[[Color_label]])
fviz_pca_biplot(result_pca,
             axes = c(2, 3),
             col.ind = data_PCA_all_vars[[Color_label]])
fviz_pca_biplot(result_pca,
             axes = c(3, 4),
             col.ind = data_PCA_all_vars[[Color_label]])

### Plot all four on one plot
# Generate the plots
p1 <- fviz_pca_biplot(result_pca, axes = c(1, 2), col.ind = data_PCA_all_vars[[Color_label]])
p2 <- fviz_pca_biplot(result_pca, axes = c(1, 3), col.ind = data_PCA_all_vars[[Color_label]])
p3 <- fviz_pca_biplot(result_pca, axes = c(2, 3), col.ind = data_PCA_all_vars[[Color_label]])
p4 <- fviz_pca_biplot(result_pca, axes = c(3, 4), col.ind = data_PCA_all_vars[[Color_label]])

# Arrange the plots in a grid
grid.arrange(p1, p2, p3, p4, nrow = 2, ncol = 2)
```


# K-Means Clustering
### 
```{r}
# Create Clusters For Four Groups
first_kmeans <- data_PCA %>% 
  kmeans(centers = 4)

# Within Group Inertia
first_kmeans$tot.withinss

kmeans_5groups<- data_PCA %>%
  kmeans(centers = 5)

kmeans_4groups<- data_PCA %>%
  kmeans(centers = 4)

kmeans_3groups<- data_PCA %>%
  kmeans(centers=3)

kmeans_2groups <- data_PCA %>%
  kmeans(centers=2)

# Extract the Clustering Groups
clusters_5groups <- kmeans_5groups$cluster
clusters_4groups <- kmeans_4groups$cluster
clusters_3groups <- kmeans_3groups$cluster
clusters_2groups <- kmeans_2groups$cluster
```

## Plot Clusters 
```{r}
# Representation in the first principal plane
fviz_pca_biplot(result_pca,
             axes = c(1, 2),
             col.ind = clusters_2groups)
fviz_pca_biplot(result_pca,
             axes = c(1, 3),
             col.ind = clusters_2groups)
fviz_pca_biplot(result_pca,
             axes = c(2, 3),
             col.ind = clusters_2groups)
fviz_pca_biplot(result_pca,
             axes = c(3, 4),
             col.ind = clusters_2groups)
```

### Comparison of 2 Groups and Distribution
```{r}
# Save each of the plots 
p1A <- fviz_pca_biplot(result_pca, axes = c(1, 2), col.ind = as.factor(clusters_2groups), col.var = "black", title = "Coloration from K-Means Clustering")
p1B <- fviz_pca_biplot(result_pca, axes = c(1, 2), col.ind = data_PCA_all_vars[[Color_label]], col.var = "black", title = "Coloration from Distribution")
p2A <- fviz_pca_biplot(result_pca, axes = c(1, 3), col.ind = as.factor(clusters_2groups), col.var = "black", title = "Coloration from K-Means Clustering")
p2B <- fviz_pca_biplot(result_pca, axes = c(1, 3), col.ind = data_PCA_all_vars[[Color_label]], col.var = "black", title = "Coloration from Distribution")
p3A <- fviz_pca_biplot(result_pca, axes = c(2, 3), col.ind = as.factor(clusters_2groups), col.var = "black", title = "Coloration from K-Means Clustering")
p3B <- fviz_pca_biplot(result_pca, axes = c(2, 3), col.ind = data_PCA_all_vars[[Color_label]], col.var = "black", title = "Coloration from Distribution")
p4A <- fviz_pca_biplot(result_pca, axes = c(3, 4), col.ind = as.factor(clusters_2groups), col.var = "black", title = "Coloration from K-Means Clustering")
p4B <- fviz_pca_biplot(result_pca, axes = c(3, 4), col.ind = data_PCA_all_vars[[Color_label]], col.var = "black", title = "Coloration from Distribution")

# Plot each axis combination
grid.arrange(p1A, p1B, nrow = 1, ncol = 2)
grid.arrange(p2A, p2B, nrow = 1, ncol = 2)
grid.arrange(p3A, p3B, nrow = 1, ncol = 2)
grid.arrange(p4A, p4B, nrow = 1, ncol = 2)
```

## Correspondance Analysis
### Her version
```{r}
# Read data
data_counts <- read.table("donnees_comptage_genre.txt",
                      sep = ";", 
                      header = TRUE, 
                      row.names = 1)
# Convert to long tibble
data_counts_long <- data_counts %>% 
  rownames_to_column(var = "Zone") %>% # Create Zone column from row names
  # Now we transform into length by aggregating all columns except zone into
  # 2 columns, one giving the Genus name, one giving the number of individuals
  pivot_longer(cols = -c("Zone"), # Variable not affected by aggregation
               names_to = "Genus", # Old column names will be 
               # brought into a "Genus" variable
               values_to = "NbIndividuals") # Numbers are aggregated in a new
# column NbIndiviuds
dim(data_counts) # Old dimensions
data_counts_long
```

### My version - Create Count of Each  
```{r}
# Select all four variables to be used in the CA and CCA, omit NAs
data_CCA <- data %>%
  # Select only the variables to use for the CA and CCA
  select(any_of(variables_to_use_CCA)) %>%
  # Exclude NAs
  na.omit() 

# Make a frequency table containing just 2 categorical variables for the CA
# With each pair of levels on a seperate row
freq_CA_long <- data_CCA %>%
  select(Family, Prefered.Habitat.Type) %>%
  table() %>%
  as_tibble %>%
  rename(Freq = "n",
         Habitat = "Prefered.Habitat.Type")

# Make a tibble with the count of each combination
freq_CA_short <- data_CCA %>%
  select(Prefered.Habitat.Type, Family) %>%
  count(Prefered.Habitat.Type, Family) %>%  
  pivot_wider(
    names_from = Family, 
    values_from = n,  
    values_fill = 0)

# Convert to dataframe and create rownames (for use creating average profile)
freq_CA_short_df <- freq_CA_short %>%
  as.data.frame() %>%  
  column_to_rownames(var = "Prefered.Habitat.Type")  

freq_CA_long
freq_CA_short
freq_CA_short_df
```

### Represent in ggplot
```{r}
ggplot(freq_CA_long) +
  aes(x = Family, y = Habitat, fill = Freq) + # fill says which column is used 
  # for filling
  geom_raster() + # represents in raster mode
  # and dress to make it pretty
  scale_fill_viridis_c() + # Change color scale
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```
### Calculate the Average Profile
```{r}
# Create average profile  
average_profile <- colSums(freq_CA_short_df) / sum(freq_CA_short_df)
average_profile

# Create the Count  
count_habitat <- rowSums(freq_CA_short_df)
count_habitat

# Division Counts
CA_repartition <- (freq_CA_short_df / count_habitat) %>% 
  rbind(Average_profile = average_profile)
CA_repartition

ggplot(freq_CA_long) +
  aes(x = Family, y = Habitat, fill = Freq) + # fill says which column is used 
  # for filling
  geom_raster() + # represents in raster mode
  # and dress to make it pretty
  scale_fill_viridis_c() + # Change color scale
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```
### CA
```{r}
# 
unique(data$Family)
result_CA <- CA(freq_CA_short_df, graph = FALSE)
fviz_eig(result_CA)
fviz_ca_biplot(result_CA, repel = TRUE)
```

## Canonical Correspondance Analysis
### Create Data
```{r}
mean_latitude_df <- data_CCA %>%
  group_by(Prefered.Habitat.Type) %>%           
  summarise(mean_latitude_abs = mean(abs(Latitude), na.rm = TRUE)) %>% 
  as.data.frame()   

mean_latitude_df
```

### Run CCA
```{r}
# Make a frequency table containing all 4 categorical variables to be used
result_CCA <- cca(freq_CA_short_df ~ mean_latitude_abs,
                    data = mean_latitude_df)
plot(result_CCA)

scores_CCA = scores(result_CCA)
ordiArrowMul(scores_CCA$biplot)
anova(result_CCA)
anova(result_CCA,by="margin")
```
### 

# Unsupervised Learning
##Split Data
###Filter Data Variables
```{r}
data_RF <- data_clean %>%
  select(any_of(variables_to_use_RF))
str(data_RF)
```

### Initial Split 
```{r}
split_data <- data_RF %>%
  initial_split(prop = 0.8, strata = "Distribution")

# Train and Test Data
train_data <- analysis(split_data)
test_data <- assessment(split_data)
```

## Random Forests
### Create the RF
```{r}
forest <- randomForest(Distribution ~ ., # Formula for prediction
                      data = train_data, # Data for training
                      ntree = 5000, # Number of trees
                      maxnodes = 10, # Number of maximum leaves for each tree
                      mtry = 4, # Number of variables for each tree
                      importance = TRUE) # Computation of importance
```

### Test the Accuracy
```{r}
# Fast way of computing accuracy with the pipe operator %>% 
predict(forest, newdata = select(test_data, - Distribution)) %>% 
  table(prediction = ., truth = test_data$Distribution) %>% 
  {sum(diag(.)) / sum(.)}
```

### Find the Variable Importance
```{r}
varImpPlot(forest)
```
### Hyper Parameter Tuning
```{r}
# Initialisation
params_CV <- trainControl(method = "repeatedcv", 
                          number = 10, 
                          repeats = 5)

predictor_forest <- caret::train(Distribution ~ ., 
                                 data = data_RF, 
                                 method = "rf", 
                                 metric = "Accuracy", 
                                 trControl = params_CV, 
                                 tuneGrid = expand.grid(mtry = 2:12))

as.data.frame(predictor_forest$results) %>% 
  ggplot(aes(x = mtry, y = Accuracy)) +
  geom_point()
```


